Semi-Permeable Vesicles: Mathematical Formulation, Numerics, and Physics

### 2020-11-25 vesicle_code/docs/permeableVes/main.tex
fixed tikzexternalize
doing shear runs for PNAS

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Quiescent Flow} 
We consider a semi-permeable elliptical vesicle with initial reduced
area $\nu = 0.65$ and several different permeability rates in a
quiescent flow. Based on our analysis, the steady state shape is a
circle with constant tension $\Lambda = 2\pi^2/L^2$.
By~\eqref{eqn:areaROC}, the area change is proportional to $\beta$, and
we expect both the time for the vesicle to deviate from its initial
area, and the time to reach its steady state to be proportional to
$\beta^{-1}$. In Table~\ref{tbl:ellipseRelaxTimes}, we report the time,
$t_\mathrm{steady}$, required for the vesicle's area to be within 1.3\%
of its maximum area, and the time, $t_\mathrm{inflate}$, required for
the vesicle's area to exceed 1.3\% of its initial area. Both times are
proportional to $\beta^{-1}$. We plot the area of a vesicle with six
different permeability rates in Figure~\ref{fig:ellipseArea}. The area
is computed directly from the vesicle shape and the time axis is scaled
by $\beta$. The plot verifies that the time for the vesicle to reach a
circular shape is proportional to $\beta^{-1}$. We also
solve~\eqref{eqn:areaROC} for the $\beta=10$ simulation and superimpose
its solution with dots on Figure~\ref{fig:ellipseArea}.
\begin{table}[htp]
  \centering
  \caption{\label{tbl:ellipseRelaxTimes} The time for a semi-permeable
  vesicle to its steady state ($t_\mathrm{steady}$) and to begin
  inflating ($t_\mathrm{inflate}$).}
  \begin{tabular}{C{0.8cm}C{2cm}C{0.8cm}C{2cm}C{0.8cm}}
    $\beta$ & $t_\mathrm{steady}$ & ratio & $t_\mathrm{inflate}$ & ratio \\
    \midrule
    $10^{0}$  & $6.45 \times 10^{0}$ & ---  
              & $2.17 \times 10^{-2}$ & --- \\ 
    $10^{-1}$ & $2.00 \times 10^{1}$ & 3.10 & 
                $1.25 \times 10^{-1}$ & 5.76 \\
    $10^{-2}$ & $9.61 \times 10^{1}$ & 4.81 & 
                $1.59 \times 10^{0}$ & 12.72 \\
    $10^{-3}$ & $8.36 \times 10^{2}$ & 8.71 & 
                $2.19 \times 10^{1}$ & 13.77 \\
    $10^{-4}$ & $8.15 \times 10^{3}$ & 9.74 & 
                $2.14 \times 10^{2}$ & 9.77 \\
    $10^{-5}$ & $8.03 \times 10^{4}$ & 9.85 & 
                $2.20 \times 10^{3}$ & 10.14 \\
    \bottomrule
  \end{tabular}
\end{table}

\begin{figure}[htp]
  \centering
%  \ifTikz
%  \input{figures/ellipseArea.tikz}
%  \fi
  \includegraphics[width=0.9\linewidth]{figures/ellipseArea}
  \caption{\label{fig:ellipseArea} The area of an elliptical
  semi-permeable vesicle, with permeability rate $\beta$ in a quiescent
  flow. The time axis is scaled by the permeability rate $\beta$. Since
  the curves collapse to a single curve, the permeability rate $\beta$
  sets a relaxation time scale.}
\end{figure}

We further analyze the transient dynamics by considering the vesicle
velocity~\eqref{eqn:vesVelocity} for small $\beta$. Two terms contribute
to the vesicle velocity: (i) the permeability velocity, $\beta(\ff \cdot
\nn)\nn$, and (ii) the force balance velocity, $\SS[\ff]$. Since $\beta
\ll 1$, the permeability velocity is much smaller than the force balance
velocity at early times, and the semi-permeable vesicle dynamics
resembles those of an impermeable vesicle. As the force balance velocity
decreases, the permeability velocity becomes significant, and the
vesicle begins to inflate to a circle. This results in the membrane
force and velocity both converging to zero. Letting $\beta = 10^{-3}$,
the normal component of the force balance velocity (blue) and the
permeability velocity (red) are plotted at three times in
Figure~\ref{fig:vesVelocity}. We also plot the velocity (green) of an
identically initialized impermeable vesicle. At the early time $t=0.5$,
the permeability velocity is negligible relative to the force balance
velocity, and the force balance velocity matches with the impermeable
vesicle velocity. At intermediate time $t=5$, all three velocities are
comparable in size, and the semi-permeable vesicle begins to inflate.
Finally, at time $t=50$, the impermeable vesicle is near steady state,
while the semi-permeable vesicle is still inflating. The right-most
plot of Figure~\ref{fig:vesVelocity} is the maximum norm of the
different velocity contributions as a function of time, and the black
vertical lines correspond to the times reported in the other three
subplots.

\begin{figure*}[htp]
  \centering
%  \input{figures/ellipseVelocity1.tikz}
%  \input{figures/ellipseVelocity2.tikz}
%  \input{figures/ellipseVelocity3.tikz}
%  \input{figures/ellipseVelocityNorms.tikz}
  \includegraphics[width=0.24\linewidth]{figures/ellipseVelocity1.pdf}
  \includegraphics[width=0.24\linewidth]{figures/ellipseVelocity2.pdf}
  \includegraphics[width=0.24\linewidth]{figures/ellipseVelocity3.pdf}
  \includegraphics[width=0.24\linewidth]{figures/ellipseVelocityNorms.pdf}
  \caption{\label{fig:vesVelocity} The velocity contributions due to
  force balance (blue) and permeability (red). The green line is the
  velocity of an impermeable vesicle with the same initial shape as the
  semi-permeable vesicle. The right-most plot is the maximum norm of
  each velocity contribution. The black lines correspond to the times
  reported in the other three plots.}
\end{figure*}

We next investigate the interplay between the tension, transmembrane
flux, and curvature of a semi-permeable vesicle. We let $\beta = 1$ and
initialize the vesicle with regions of both positive and negative
curvature. The vesicle tension, $\Lambda$, and the flux, $\ff \cdot
\nn$, are plotted in Figure~\ref{fig:starTensionFlux}. At early times,
regions of large negative curvature have negative tension and regions
with large positive curvature have positive tension, and the same result
holds for an impermeable vesicle. In addition, regions with large
curvature have a large inflow of fluid (positive flux), while the
flatter regions have an outflow of fluid. At these early times, the net
effect is vesicle deflation. As the vesicle shape relaxes, the same
general tension and flux trends persist, but the net flux is positive
and the vesicle inflates. At later times, the tension becomes positive
and the vesicle continues to inflate towards a circle. Finally, the
tension transitions to a positive constant function, the membrane force
and flux approach zero, and the vesicle shape reaches its steady
circular shape.

@@

\begin{figure*}[htp]
  \centering
  \includegraphics[width=0.16\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarTensionTime1.pdf}
  \includegraphics[width=0.16\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarTensionTime2.pdf}
  \includegraphics[width=0.16\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarTensionTime3.pdf}
  \includegraphics[width=0.16\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarTensionTime4.pdf}
  \includegraphics[width=0.16\linewidth,trim =2cm 5cm 0cm 5cm,clip=true]{figures/StarTensionTime5.pdf}
  \includegraphics[width=0.16\linewidth,trim =2cm 5cm 0cm 5cm,clip=true]{figures/StarTensionTime6.pdf}

  \includegraphics[width=0.16\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarFluxTime1.pdf}
  \includegraphics[width=0.16\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarFluxTime2.pdf}
  \includegraphics[width=0.16\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarFluxTime3.pdf}
  \includegraphics[width=0.16\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarFluxTime4.pdf}
  \includegraphics[width=0.16\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarFluxTime5.pdf}
  \includegraphics[width=0.16\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarFluxTime6.pdf}
  \caption{\label{fig:starTensionFlux} The vesicle tension (top) and
  flux (bottom) of the semi-permeable vesicle in a quiescent
  flow with $\beta=1$.}
\end{figure*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Shear Flow}
We consider a semi-permeable vesicle in the shear flow
$\uu_{\infty}(\xx) = \dot{\gamma} (y,0)$ with different permeability
rates $\beta$ and shear rates $\dot{\gamma}$. Depending on its reduced
area and viscosity contrast, two-dimensional vesicles either tank tread
or tumble~\cite{fin-lam-sei-gom2008,kra-win-sei-lip1996}. Since we are
considering semi-permeable vesicles, we assume there is no viscosity
contrast, and tumbling is not possible.
\begin{figure}[htp]
  \centering
%  \ifTikz
%  \input{figures/shearSS.tikz}
%  \fi
  \includegraphics[width=0.9\linewidth]{figures/shearSS.pdf}
  \caption{\label{fig:shearSS} The reduced area of a semi-permeable
  vesicle in a shear flow. The final reduced area is nearly independent
  of the initial reduced area.}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Quiescent Flow} 
Based on our analysis, a semi-permeable vesicle in a quiescent flow
relaxes to a circle with constant tension $\Lambda = 2\pi^2/L^2$.
Furthermore, equation~\eqref{eqn:areaROC} shows that the area change is
proportional to $\beta$, so we expect the times for the vesicle to
deviate from its initial reduced area and to reach its steady state to
both be proportional to $\beta^{-1}$. This is verified by considering a
vesicle with initial reduced area $\nu = 0.65$ and several different
semi-permeability rates. In Table~\ref{tbl:ellipseRelaxTimes} we report
the time, $t_\mathrm{steady}$, required for the vesicle's area to be
within 1.3\% of its steady state reduced area, and the time,
$t_\mathrm{inflate}$, required for the vesicle's reduced area to exceed
1.3\% of its initial reduced area. Both times are proportional to
$\beta^{-1}$. 
\begin{table}[htp]
  \centering
  \caption{\label{tbl:ellipseRelaxTimes} The time for a semi-permeable
  vesicle to its steady state ($t_\mathrm{steady}$) and to begin
  inflating ($t_\mathrm{inflate}$).}
  \begin{tabular}{C{0.8cm}C{2cm}C{0.8cm}C{2cm}C{0.8cm}}
    $\beta$ & $t_\mathrm{steady}$  & ratio 
            & $t_\mathrm{inflate}$ & ratio \\
    \midrule
    $10^{0}$  & $6.45 \times 10^{0}$ & ---  
              & $2.17 \times 10^{-2}$ & --- \\ 
    $10^{-1}$ & $2.00 \times 10^{1}$ & 3.10 & 
                $1.25 \times 10^{-1}$ & 5.76 \\
    $10^{-2}$ & $9.61 \times 10^{1}$ & 4.81 & 
                $1.59 \times 10^{0}$ & 12.72 \\
    $10^{-3}$ & $8.36 \times 10^{2}$ & 8.71 & 
                $2.19 \times 10^{1}$ & 13.77 \\
    $10^{-4}$ & $8.15 \times 10^{3}$ & 9.74 & 
                $2.14 \times 10^{2}$ & 9.77 \\
    $10^{-5}$ & $8.03 \times 10^{4}$ & 9.85 & 
                $2.20 \times 10^{3}$ & 10.14 \\
    \bottomrule
  \end{tabular}
\end{table}

In Figure~\ref{fig:relaxationArea}, we examine the transient dynamics of
four vesicles $\beta = 10^{-3}$ and a variety of initial configurations.
First, in all cases the vesicle eventually reaches a circular shape of
reduced area $\nu=1$. Second, we observe that even though the vesicle is
inflating to a circle, meaning that there is a net flux into the
vesicle, there are regions where the flux is out of the vesicle. These
regions are denoted in blue and are only present at early times.
Finally, the vesicles area does not initially change by much, and
intermediate shapes of semi-permeable vesicles resemble the shape of an
impermeable vesicle. This can be explained by considering the two terms
that contribute to the vesicle velocity: (i) the permeability velocity,
$\beta(\ff \cdot \nn) \nn$, and (ii) the force balance velocity,
$\SS[\ff]$. Since $\beta = 10^{-3}$, the permeability velocity is much
smaller than the force balance velocity at early times, and the
semi-permeable vesicle dynamics resembles those of an impermeable
vesicle. As the force balance velocity decreases, the permeability
velocity dominates, and the vesicle begins to inflate to a circle with
no vanishing force and velocity.

\begin{figure*}[htp]
  \centering
  \input{figures/relaxationArea.tikz}
  \caption{\label{fig:relaxationArea} (a): Snapshots of four
  semi-permeable vesicles with $\beta = 10^{-3}$ submerged in a
  quiescent flow. The fluid flux along the red regions is into the
  vesicle, and the fluid flux along the blue regions is out of the
  vesicle. In all cases, the vesicle reaches a steady-state circular
  shape. (b): The reduced area of each of the simulations. The
  particular snapshots in (a) occur at the marks along the curve.}
\end{figure*}

%We plot the area of a vesicle with six different permeability rates in
%Figure~\ref{fig:ellipseArea}. The area is computed directly from the
%vesicle shape and the time axis is scaled by $\beta$. The plot verifies
%that the time for the vesicle to reach a circular shape is proportional
%to $\beta^{-1}$. We also solve~\eqref{eqn:areaROC} for the $\beta=10$
%simulation and superimpose its solution with dots on
%Figure~\ref{fig:ellipseArea}.

%\begin{figure}[htp]
%  \centering
%%  \ifTikz
%%  \input{figures/ellipseArea.tikz}
%%  \fi
%  \includegraphics[width=0.9\linewidth]{figures/ellipseArea}
%  \caption{\label{fig:ellipseArea} The area of an elliptical
%  semi-permeable vesicle, with permeability rate $\beta$ in a quiescent
%  flow. The time axis is scaled by the permeability rate $\beta$. Since
%  the curves collapse to a single curve, the permeability rate $\beta$
%  sets a relaxation time scale.}
%\end{figure}

%We further analyze the transient dynamics by considering the vesicle
%velocity~\eqref{eqn:vesVelocity} for small $\beta$. Two terms contribute
%to the vesicle velocity: (i) the permeability velocity, $\beta(\ff \cdot
%\nn)\nn$, and (ii) the force balance velocity, $\SS[\ff]$. Since $\beta
%\ll 1$, the permeability velocity is much smaller than the force balance
%velocity at early times, and the semi-permeable vesicle dynamics
%resembles those of an impermeable vesicle. As the force balance velocity
%decreases, the permeability velocity becomes significant, and the
%vesicle begins to inflate to a circle. This results in the membrane
%force and velocity both converging to zero. Letting $\beta = 10^{-3}$,
%the normal component of the force balance velocity (blue) and the
%permeability velocity (red) are plotted at three times in
%Figure~\ref{fig:vesVelocity}. We also plot the velocity (green) of an
%identically initialized impermeable vesicle. At the early time $t=0.5$,
%the permeability velocity is negligible relative to the force balance
%velocity, and the force balance velocity matches with the impermeable
%vesicle velocity. At intermediate time $t=5$, all three velocities are
%comparable in size, and the semi-permeable vesicle begins to inflate.
%Finally, at time $t=50$, the impermeable vesicle is near steady state,
%while the semi-permeable vesicle is still inflating. The right-most
%plot of Figure~\ref{fig:vesVelocity} is the maximum norm of the
%different velocity contributions as a function of time, and the black
%vertical lines correspond to the times reported in the other three
%subplots.
%
%\begin{figure*}[htp]
%  \centering
%%  \scalebox{0.6}{\input{figures/ellipseVelocity1.tikz}}
%%  \scalebox{0.6}{\input{figures/ellipseVelocity2.tikz}}
%%  \scalebox{0.6}{\input{figures/ellipseVelocity3.tikz}}
%%  \scalebox{0.6}{\input{figures/ellipseVelocity4.tikz}}
%%  \input{figures/ellipseVelocityNorms.tikz}
%  \includegraphics[width=0.24\linewidth]{figures/ellipseVelocity1.pdf}
%  \includegraphics[width=0.24\linewidth]{figures/ellipseVelocity2.pdf}
%  \includegraphics[width=0.24\linewidth]{figures/ellipseVelocity3.pdf}
%  \includegraphics[width=0.24\linewidth]{figures/ellipseVelocity4.pdf}
%  \caption{\label{fig:vesVelocity} The normal velocity contributions due
%  to force balance (blue) and permeability (red). The green line is the
%  velocity of an impermeable vesicle with the same initial shape as the
%  semi-permeable vesicle.}
%\end{figure*}

%We next investigate the interplay between the tension, transmembrane
%flux, and curvature of a semi-permeable vesicle. We let $\beta = 1$ and
%initialize the vesicle with regions of both positive and negative
%curvature. The vesicle tension, $\Lambda$, and the flux, $\ff \cdot
%\nn$, are plotted in Figure~\ref{fig:starTensionFlux}. At early times,
%regions of large negative curvature have negative tension and regions
%with large positive curvature have positive tension, and the same result
%holds for an impermeable vesicle. In addition, regions with large
%curvature have a large inflow of fluid (positive flux), while the
%flatter regions have an outflow of fluid. At these early times, the net
%effect is vesicle deflation. As the vesicle shape relaxes, the same
%general tension and flux trends persist, but the net flux is positive
%and the vesicle inflates. At later times, the tension becomes positive
%and the vesicle continues to inflate towards a circle. Finally, the
%tension transitions to a positive constant function, the membrane force
%and flux approach zero, and the vesicle shape reaches its steady
%circular shape.

@@

%\begin{figure*}[htp]
%  \centering
%  \includegraphics[width=0.16\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarTensionTime1.pdf}
%  \includegraphics[width=0.16\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarTensionTime2.pdf}
%  \includegraphics[width=0.16\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarTensionTime3.pdf}
%  \includegraphics[width=0.16\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarTensionTime4.pdf}
%  \includegraphics[width=0.16\linewidth,trim =2cm 5cm 0cm 5cm,clip=true]{figures/StarTensionTime5.pdf}
%  \includegraphics[width=0.16\linewidth,trim =2cm 5cm 0cm 5cm,clip=true]{figures/StarTensionTime6.pdf}
%
%  \includegraphics[width=0.16\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarFluxTime1.pdf}
%  \includegraphics[width=0.16\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarFluxTime2.pdf}
%  \includegraphics[width=0.16\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarFluxTime3.pdf}
%  \includegraphics[width=0.16\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarFluxTime4.pdf}
%  \includegraphics[width=0.16\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarFluxTime5.pdf}
%  \includegraphics[width=0.16\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarFluxTime6.pdf}
%  \caption{\label{fig:starTensionFlux} The vesicle tension (top) and
%  flux (bottom) of the semi-permeable vesicle in a quiescent
%  flow with $\beta=1$.}
%\end{figure*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Shear Flow}
We consider a semi-permeable vesicle in the shear flow
$\uu_{\infty}(\xx) = \dot{\gamma} (y,0)$ with different permeability
rates $\beta$ and shear rates $\dot{\gamma}$. Depending on its reduced
area and viscosity contrast, two-dimensional vesicles either tank tread
or tumble~\cite{fin-lam-sei-gom2008,kra-win-sei-lip1996}. Since we are
considering semi-permeable vesicles, we assume there is no viscosity
contrast, and tumbling is not possible.
\begin{figure}[htp]
  \centering
  \ifTikz
  \input{figures/shearSS.tikz}
  \fi
%  \includegraphics[width=0.9\linewidth]{figures/shearSS.pdf}
  \caption{\label{fig:shearSS} The reduced area of a semi-permeable
  vesicle in a shear flow. The final reduced area is nearly independent
  of the initial reduced area.}
\end{figure}
